---
title: "Biomarker Analysis"
author: "Munir Winkel, PhD "
date: "December 2021, amended in February, June and August 2022 by and Pascale Vonaesch, PhD"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: tango
    df_print: paged
  html_notebook:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r global options, include=FALSE, warning=FALSE, message=FALSE, echo= FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE
                      , error = FALSE, eval = TRUE )
#-- 3d plot option
options(rgl.useNULL = TRUE) 
library(magrittr); library(ggplot2); library(rgl); library(dplyr); library(vtable); library(flextable)

rm(list = ls())
setwd( "/Users/admin/Desktop/Papers and manuscripts/under revision/Biomarker paper/Revisions/Ranalysis")
source("useful_functions.R") 
devAskNewPage(ask = FALSE)

# data to be taken out as we have multiple measurements or as they are below the detection limit, i added back the igas and the gmu and took out all cytokines in feces and duodenum
cytokinesbelowdetection <- c('gcsf_feces', 'gmcsf_feces','il13_feces','il15_feces','il17_feces','il1b_duo','il1b_feces','il2r_feces','il4_feces','il5_feces','il6_feces','il7_feces','il8_feces','mcp1_feces','mig_feces','mip1a_feces','mip1b_feces','rantes_duo','tnfa_feces','egf_feces','hgf_feces','il10_feces','eotaxin_feces','fgfbasic_feces','ifna_feces','ifng_feces','il1ra_feces','il4_duo','il8_duo','il2r_duo','ifna_duo','il12_feces','il1ra_duo','il2_feces','il6_duo','il7_duo','mig_duo','mcp1_duo','vegf_feces','vegf_duo','ip10_feces','il2_duo','il5_duo','ip10_duo','rantes_feces','tnfa_duo','mip1a_duo','mip1b_duo')



#-- custom dataset
datt = read.csv("/Users/admin/Desktop/Papers and manuscripts/under revision/Biomarker paper/Revisions/Supplements/Supplementary File 1_biomarker_data_revised.csv" ) %>%
  dplyr::select( - cytokinesbelowdetection) %>%
  tibble::as_tibble()

# look at the variables that are in the dataset
colnames(datt)
nrow(datt)
table(datt$pays)
table(datt$pays, datt$stunted)

# how many samples are shared in between the different biomarkers? il10 blood as a proxy for all blood cytokines (missing imputed), iga_feces as a proxy for gut Immunoglobulins, bcca_geomean as a proxy of branched chain amino acids, gmu200 as a proxy for endocab values


controls = c( "citrulline","iga_feces","AAT_feces", "iga_blood"
              ,"il10_blood", "crp", "cal_feces", "valin", "gmu200", "igf_n")
datt_biomarkers<-datt[, c(controls)]

sumtable(datt_biomarkers,
         summ=c('notNA(x)',
                'propNA(x)'))

datt_Mada<-subset(datt, pays=="Madagascar")
datt_CAR<-subset(datt, pays=="RCA")
datt_Mada_bm<-datt_Mada[, c(controls)]
datt_CAR_bm<-datt_CAR[, c(controls)]

sumtable(datt_Mada_bm,
         summ=c('notNA(x)',
                'propNA(x)'))

sumtable(datt_CAR_bm,
         summ=c('notNA(x)',
                'propNA(x)'))

calpro<-datt[which(!is.na(datt$cal_feces)==TRUE), ]
aat<-datt[which(!is.na(datt$AAT_feces)==TRUE), ]
citrullin<-datt[which(!is.na(datt$citrulline)==TRUE), ]
igfeces<-datt[which(!is.na(datt$iga_feces)==TRUE), ]
igblood<-datt[which(!is.na(datt$iga_blood)==TRUE), ]
cyto<-datt[which(!is.na(datt$il10_blood)==TRUE), ]
endocab<-datt[which(!is.na(datt$gmu200)==TRUE), ]
crp<-datt[which(!is.na(datt$crp)==TRUE), ]
igf<-datt[which(!is.na(datt$igf_n)==TRUE), ]
branchedAA<-datt[which(!is.na(datt$bcaa_geomean)==TRUE), ]


nrow(intersect(calpro, aat))
nrow(intersect(calpro, igfeces))
nrow(intersect(calpro, igblood))
nrow(intersect(calpro, cyto))
nrow(intersect(calpro, citrullin))
nrow(intersect(calpro, crp))
nrow(intersect(calpro, endocab))
nrow(intersect(calpro, igf))
nrow(intersect(calpro, branchedAA))

nrow(intersect(aat, igfeces))
nrow(intersect(aat, igblood))
nrow(intersect(aat, cyto))
nrow(intersect(aat, citrullin))
nrow(intersect(aat, crp))
nrow(intersect(aat, endocab))
nrow(intersect(aat, igf))
nrow(intersect(aat, branchedAA))

nrow(intersect(igfeces, igblood))
nrow(intersect(igfeces, cyto))
nrow(intersect(igfeces, citrullin))
nrow(intersect(igfeces, crp))
nrow(intersect(igfeces, endocab))
nrow(intersect(igfeces, igf))
nrow(intersect(igfeces, branchedAA))

nrow(intersect(igblood, cyto))
nrow(intersect(igblood, citrullin))
nrow(intersect(igblood, crp))
nrow(intersect(igblood, endocab))
nrow(intersect(igblood, igf))
nrow(intersect(igblood, branchedAA))

nrow(intersect(cyto, citrullin))
nrow(intersect(cyto, crp))
nrow(intersect(cyto, endocab))
nrow(intersect(cyto, igf))
nrow(intersect(cyto, branchedAA))

nrow(intersect(citrullin, crp))
nrow(intersect(citrullin, endocab))
nrow(intersect(citrullin, igf))
nrow(intersect(citrullin, branchedAA))

nrow(intersect(crp, endocab))
nrow(intersect(crp, igf))
nrow(intersect(crp, branchedAA))

nrow(intersect(endocab, igf))
nrow(intersect(endocab, branchedAA))

nrow(intersect(igf, branchedAA))



```

## Description of Statistical Methods

**Linear Regression** 
We model continuous response variables (such as height-for-age-z-scores) as a linear function of variables of interest and control variables (such as country of origin, sex, and age). In some cases, we take a square-root or logarithm of the response variable so that the residual terms are closer to being normally distributed. When sample sizes are small, we regress the outcome variable on the first two principle components of a set of variables (e.g. the endocab variables). 

**Bootstrapping**
Sometimes there are potential outliers in a dataset that have a large influence on the parameter estimates from the linear models. One way of making the analysis more robust is to take samples of the data (with replacement) and analyze the data multiple times. This bootstrapped approach allows us to see which results are sensitive to outliers and which ones are robust. 


## Overview 


```{r}
# --- recode variables that can be reported as categorical variables and are not yet categorized

# generate a new variable ferritine and ferritine seuil which is ajusted for inflammation

# generate a new variable for ferritine levels

datt$valine_level<-as.factor(datt$valine_level)
levels(datt$valine_level) <- c('low', 'normal', 'elevated')
table(datt$valine_level)

datt$leucine_level<-as.factor(datt$leucine_level)
levels(datt$leucine_level) <- c('low', 'normal', 'elevated')
table(datt$leucine_level)

datt$isoleucine_level<-as.factor(datt$isoleucine_level)
levels(datt$isoleucine_level) <- c('low', 'normal', 'elevated')
table(datt$isoleucine_level)

datt$alanine_level<-as.factor(datt$alanine_level)
levels(datt$alanine_level) <- c('low', 'normal', 'elevated')
table(datt$alanine_level)

datt$bcaa_geomean<-as.numeric(datt$bcaa_geomean)


datt <- datt %>% 
  dplyr::mutate( 
    # ferritine 
    ferritine_seuil = dplyr::case_when(
        ferritine < 11  ~ "Yes",
        ferritine >= 11 ~ "Non",
        is.na( ferritine) ~ "Missing")
    # calpro 
  , calprotectinlevel = dplyr::case_when(
      cal_feces_wet >150  ~ "elevated",
      cal_feces_wet >100 & age > 36  ~ "elevated", 
      cal_feces_wet <150 & age < 36  ~ "normal",
      cal_feces_wet < 100 ~ "normal")
   # aat
  , AATlevel = dplyr::case_when(
      AAT_feces > 2  ~ "elevated",
      AAT_feces < 1.25 | AAT_feces_wet < 0.15   ~ "normal",
      AAT_feces < 2   ~ "grey zone", 
      is.na(AAT_feces) ~ "missing")
  # citrulline
  , citrullinelevel = dplyr::case_when(
      citrulline < 16 & age <= 36 & age >24  ~ "too low",
      citrulline >= 16 & age <= 36 & age >24  ~ "normal",
      citrulline > 40 & age <= 36 & age >24  ~ "too high",
      citrulline < 7 & age > 36 | age <=24  ~ "too low",
      citrulline >= 7 & age > 36 | age <=24  ~ "normal",
      citrulline > 43 & age > 36 | age <=24  ~ "too high",
      is.na(citrulline) ~ "missing" )
  # crp 
  , crp_seuil = dplyr::case_when(
      crp <= 10  ~ "normal",
      crp > 10  ~ "elevated",
      is.na(crp) ~ "missing")


    )


table(datt$citrullinelevel, datt$pays)
100/(13+382)*13
100/(29+352)*29

table(datt$crp_seuil, datt$pays)
100/(34+371)*34
100/(68+274)*68

table(datt$AATlevel, datt$pays)
100/(27+379)*27
100/(8+376)*8

table(datt$calprotectinlevel, datt$pays)
100/(138+237)*138
100/(84+270)*84

table(endocab$stunted)
t.test(gmu200 ~ stunted, data = endocab)
t.test(amu ~ stunted, data = endocab)
t.test(mmu ~ stunted, data = endocab)

t.test(bcaa_geomean ~ stunted, data = datt)



#--useful variable lists

blood_cytos = c('fgfbasic_blood', 'il1b_blood', 'gcsf_blood'
                , 'il10_blood', 'il13_blood', 'il6_blood', 'il12_blood'
                , 'rantes_blood', 'eotaxin_blood', 'il17_blood', 'mip1a_blood'
                , 'gmcsf_blood', 'mip1b_blood', 'mcp1_blood', 'il15_blood'
                , 'egf_blood', 'il5_blood', 'hgf_blood', 'vegf_blood'
                , 'ifng_blood'
                , 'ifna_blood', 'il1ra_blood', 'tnfa_blood', 'il2_blood'
                , 'il7_blood'
                , 'ip10_blood', 'il2r_blood', 'mig_blood', 'il4_blood'
                , 'il8_blood')


blood_immu = c( 'ige_blood', 'igg1_blood', 'igg2_blood', 'igg3_blood'
             , 'igg4_blood',  'igm_blood', 'iga_blood', 'fgfbasic_blood'
             ,'il1b_blood', 'gcsf_blood'
             , 'il10_blood', 'il13_blood', 'il6_blood', 'il12_blood'
             , 'rantes_blood', 'eotaxin_blood', 'il17_blood', 'mip1a_blood'
             , 'gmcsf_blood', 'mip1b_blood', 'mcp1_blood', 'il15_blood'
             , 'egf_blood', 'il5_blood', 'hgf_blood', 'vegf_blood'
             , 'ifng_blood'
             , 'ifna_blood', 'il1ra_blood', 'tnfa_blood', 'il2_blood'
             , 'il7_blood'
             , 'ip10_blood', 'il2r_blood', 'mig_blood', 'il4_blood'
             , 'il8_blood')

plates = c('plate_iga_duo', 'plate_ige_duo', 'plate_igg_igm_duo'
           , 'plate_blood', 'plate_iggam_blood', 'plate_ige_blood'
           , 'plate_feces', 'iga_plate_feces', 'igg_plate_feces' )

immus = c(  'iga_feces', 'iga_blood', 'iga_duo'
          , 'ige_blood', 'ige_duo', 'igg1_blood', 'igg1_duo', 'igg1_feces'
          , 'igg2_blood', 'igg2_duo', 'igg2_feces', 'igg3_blood', 'igg3_duo'
          , 'igg3_feces', 'igg4_blood', 'igg4_duo', 'igg4_feces', 'igm_blood'
          , 'igm_duo', 'igm_feces'
          , 'fgfbasic_blood', 'il1b_blood', 'gcsf_blood'
          , 'il10_blood', 'il13_blood', 'il6_blood', 'il12_blood'
          , 'rantes_blood', 'eotaxin_blood', 'il17_blood', 'mip1a_blood'
          , 'gmcsf_blood', 'mip1b_blood', 'mcp1_blood', 'il15_blood'
          , 'egf_blood', 'il5_blood', 'hgf_blood', 'vegf_blood', 'ifng_blood'
          , 'ifna_blood', 'il1ra_blood', 'tnfa_blood', 'il2_blood', 'il7_blood'
          , 'ip10_blood', 'il2r_blood', 'mig_blood', 'il4_blood', 'il8_blood')

immunoglob = c('iga_feces', 'iga_blood', 'iga_duo'
          , 'ige_blood', 'ige_duo', 'igg1_blood', 'igg1_duo', 'igg1_feces'
          , 'igg2_blood', 'igg2_duo', 'igg2_feces', 'igg3_blood', 'igg3_duo'
          , 'igg3_feces', 'igg4_blood', 'igg4_duo', 'igg4_feces', 'igm_blood'
          , 'igm_duo', 'igm_feces')


non_duo_immus = immus[ !grepl( "_duo", immus)]
duo_immus     = immus[  grepl( "_duo", immus)]


dat = datt %>% dplyr::select( -c( tpi_level, stunted, haz, anemia
                                  , dplyr::all_of( plates )))


#-- useful datasets
immu_dat = datt %>% dplyr::select( all_of(immus)
                      , all_of(plates), all_of(blood_cytos)
                      , pays, sex, id, haz, stunted )

#--- section of code to identify potential outliers
long = immu_dat %>%
  dplyr::select( - all_of( plates) ) %>% 
  tidyr::pivot_longer( col = dplyr::all_of( immus ) ) %>%
  dplyr::arrange( name, id ) %>%
  dplyr::group_by( name ) %>%
  dplyr::mutate( upper99 =  ifelse( is.numeric(value), quantile(value, probs = .99, na.rm = TRUE ), NA)
  )

#-- values greater than twice the 99th percentile of each variable
high = long %>% dplyr::select( - value ) %>%
  tidyr::pivot_wider(values_from = "upper99", id_cols = c(id, sex, pays ) ) %>%
  dplyr::mutate( id = "upper99")

#--- creating the potential outlier dataset
bads = long %>% 
  dplyr::filter( value > 2 * upper99 ) %>% 
  dplyr::select( - upper99) %>% 
  tidyr::pivot_wider( id_cols = c(id, sex, pays))

bad_ids = bads$id %>% unique()
#--- the values that trigger notification that it's an outlier
high = data.frame( high[1, colnames(high) %in% colnames(bads)] )

###--- visualizing the outliers in a PCoA type analysis
immu_pcoa = immu_dat %>% 
  dplyr::mutate( flagged = ( id %in% bad_ids) ) %>%
  non_missing( controls = immus[ grepl("feces", immus)]) %>%
  make_pca_vars( normer = NULL, amp_vars = immus[ grepl("feces", immus)], number_pcs = 3 )
# dim(immu_pcoa)

long = long %>%
  dplyr::mutate( flagged = ( id %in% bad_ids) )

```

First we provide an overview of the main variables, by country, and then by country and stunting status. 

```{r}


##--- table1 features for all covariables
table1<-table1::table1( data = datt, 
                ~ haz + haz_cont + age + hemoglobine + anemia +  ascaris+ trichuris+ giardiase + num_genes
                 + sda +  saison |  pays 
                , overall = FALSE)

table1<-as.data.frame(table1)

write.csv(table1, "table1.csv")

##--- table2 features for all biomarkers


table2<-table1::table1( data = datt, 
                ~  lm_ratio + lactitol + tpi_level + ferritine + ferritine_seuil
                + citrulline + citrullinelevel+ valin + valine_level + alanine + alanine_level + isoleucin + isoleucine_level + leucin + leucine_level + AAT_duo + AAT_feces + AATlevel 
                  + cal_duo + cal_feces + calprotectinlevel
                + crp + crp_seuil + igf_n 
                |  pays + stunted
                , overall = FALSE)

table2<-as.data.frame(table2)

write.csv(table2, "table2.csv")


##--- tableS1 immunoglobulins

tableS1<-table1::table1( data = datt, 
                ~  iga_feces + ige_blood+ ige_duo+ igg1_blood+ igg1_duo+ igg1_feces+ igg2_blood+ igg2_duo+ igg2_feces+ igg3_blood+ igg3_duo+ igg3_feces+ igg4_blood+ igg4_duo+ igg4_feces+ igm_blood+ igm_duo+ igm_feces 
                |  pays + stunted
                , overall = FALSE)

tableS1

tableS1<-as.data.frame(tableS1)

write.csv(tableS1, "tableS1.csv")

##--- tableS2 cytokines blood

tableS2<-table1::table1( data = datt, 
                ~  fgfbasic_blood
             +il1b_blood+ gcsf_blood
             + il10_blood+ il13_blood+ il6_blood+ il12_blood
             + rantes_blood+ eotaxin_blood+ il17_blood+ mip1a_blood
             + gmcsf_blood+ mip1b_blood+ mcp1_blood+ il15_blood
             + egf_blood+ il5_blood+ hgf_blood+ vegf_blood
             + ifng_blood
             + ifna_blood+ il1ra_blood+ tnfa_blood+ il2_blood
             + il7_blood
             + ip10_blood+ il2r_blood+ mig_blood+ il4_blood
             + il8_blood
                |  stunted
                , overall = FALSE)

tableS2

tableS2<-as.data.frame(tableS2)

write.csv(tableS2, "tableS2.csv")

##--- tableS3 endocab blood

tableS3<-table1::table1( data = datt, 
                ~  amu + mmu+ gmu200 
                |  stunted
                , overall = FALSE)

tableS3

tableS3<-as.data.frame(tableS3)

write.csv(tableS3, "tableS3.csv")
```

A large number of the variables have missing values. These variables are missing between 3% to 92% of the values, where the majority of variables are missing less than 15% of their values. The endocab variables are missing 92.4% of observations. The blood cytokine values are missing 91% of observations. Immunoglobulin variables in the duodenum are missing aroud 80% of values.

The variables igf_n, tpi_level, lm_ratio, and duodenal immunoglobulin variables are missing more than 50% of their values. Lactitol is missing 27% of its values, virulence gene count is missing around 25%, and AAT_feces is missing 18.9%. 


```{r, fig.width = 12}

## --- table of available sample sets 
length(which(!is.na(datt$plate_blood))==TRUE) # available for blood cytokines: 66
length(which(!is.na(datt$plate_ige_blood))==TRUE) # available for ige blood: 766
length(which(!is.na(datt$plate_iggam_blood))==TRUE) # available for ige blood: 769
length(which(!is.na(datt$citrulline))==TRUE) # available for citrulline: 776
length(which(!is.na(datt$crp))==TRUE) # available for crp: 747
length(which(!is.na(datt$anemia))==TRUE) # available for anemia: 748

length(which(!is.na(datt$plate_igg_igm_duo))==TRUE) # available for duodenal IgGandM: 129
length(which(!is.na(datt$plate_iga_duo))==TRUE) # available for duodenal IgA: 128
length(which(!is.na(datt$plate_ige_duo))==TRUE) # available for duodenal Ige: 130
length(which(!is.na(datt$AAT_duo))==TRUE) # available for duodenal AAT: 139
length(which(!is.na(datt$cal_duo))==TRUE) # available for duodenal calpro: 187

length(which(!is.na(datt$plate_feces))==TRUE) # available for fecal data Immunoglobulins: 715
length(which(!is.na(datt$cal_feces))==TRUE) # available for fecal calpro: 732
length(which(!is.na(datt$AAT_feces))==TRUE) # available for fecal AAT: 783

length(which(!is.na(datt$lm_ratio))==TRUE) # available for lm ratio: 265
length(which(!is.na(datt$igf_n))==TRUE) # available for igf: 364
length(which(!is.na(datt$mmu))==TRUE) # available for mmu: 61
length(which(!is.na(datt$amu))==TRUE) # available for amu: 61
length(which(!is.na(datt$gmu200))==TRUE) # available for gmu200: 61

# --- univariate analysis with stunting of main variables

t.test(igf_n ~ stunted, data = datt)
t.test(cal_feces ~ stunted, data = datt)
t.test(AAT_feces ~ stunted, data = datt)
t.test(cal_duo ~ stunted, data = datt)
t.test(cal_feces ~ stunted, data = datt)
t.test( lm_ratio~ stunted, data = datt)
t.test( citrulline~ stunted, data = datt) # p-value = 0.0009384
t.test( valin~ stunted, data = datt) # p-value = 1.085e-06
t.test( leucine~ stunted, data = datt) # p-value = 1.085e-06
t.test( isoleucin~ stunted, data = datt) # p-value = 7.947e-05
t.test( bcaa_geomean~ stunted, data = datt) # p-value = 1.383e-07
t.test( alanine~ stunted, data = datt) # p-value = ns

t.test( crp~ stunted, data = datt)

chisq.test(datt$calprotectinlevel, datt$stunted)
chisq.test(datt$AATlevel, datt$stunted)

chisq.test(datt$citrullinelevel, datt$stunted) #p-value = 0.06569
boxplot(datt$citrulline~datt$stunted)

chisq.test(datt$valine_level, datt$stunted) #p-value = 3.598e-05
boxplot(datt$valin~datt$stunted)

chisq.test(datt$isoleucine_level, datt$stunted) #p-value = 0.0001877
boxplot(datt$isoleucin~datt$stunted)

chisq.test(datt$leucine_level, datt$stunted) #p-value = 2.812e-07
boxplot(datt$leucin~datt$stunted)

boxplot(datt$bcaa_geomean~datt$stunted)

chisq.test(datt$alanine_level, datt$stunted) #p-value = 0.008

datt_stunted<-filter(datt, stunted=="stunted")
datt_nonstunted<-filter(datt, stunted=="non-stunted")

stunted<-mean(datt_stunted$bcaa_geomean, na.rm=TRUE)
nonstunted<-mean(datt_nonstunted$bcaa_geomean, na.rm=TRUE)
log2(stunted/nonstunted) #-0.1306229
stunted/nonstunted #0.91

stunted<-mean(datt_stunted$valin, na.rm=TRUE)
nonstunted<-mean(datt_nonstunted$valin, na.rm=TRUE)
log2(stunted/nonstunted) #-0.1265831

stunted<-mean(datt_stunted$leucin, na.rm=TRUE)
nonstunted<-mean(datt_nonstunted$leucin, na.rm=TRUE)
log2(stunted/nonstunted) #-0.1493881

stunted<-mean(datt_stunted$isoleucin, na.rm=TRUE)
nonstunted<-mean(datt_nonstunted$isoleucin, na.rm=TRUE)
log2(stunted/nonstunted) #-0.1119716

stunted<-mean(datt_stunted$citrulline, na.rm=TRUE)
nonstunted<-mean(datt_nonstunted$citrulline, na.rm=TRUE)
log2(stunted/nonstunted) #-0.1044675
stunted/nonstunted #0.93

chisq.test(datt$alanine_level, datt$stunted) #p-value = 0.008073
boxplot(datt$alanine~datt$stunted)

chisq.test(datt$tpi_level, datt$stunted)
chisq.test(datt$crp_seuil, datt$stunted) #p-value = 0.02464
boxplot(log(datt$crp)~datt$stunted)

t.test( iga_feces~ stunted, data = datt) # p-value = 0.7968
boxplot(log(datt$iga_feces)~datt$stunted)

t.test( iga_feces~ stunted, data = datt) # p-value = 0.7968
boxplot(log(datt$iga_feces)~datt$stunted)

t.test(AAT_feces~ stunted, data = datt) # p-value = 0.7968
boxplot(log(datt$AAT_feces)~datt$stunted)

t.test(cal_feces~ stunted, data = datt) # p-value = 0.184
boxplot(log(datt$cal_feces)~datt$stunted)

t.test(igf_n~ stunted, data = datt) # p-value = 0.3168
boxplot(log(datt$igf_n)~datt$stunted)



##--- missing data plots
g1 = finalfit::missing_plot(dat %>% dplyr::filter( pays == "RCA"), title = "Central African Republic" )
g2 = finalfit::missing_plot( dat %>% dplyr::filter( pays != "RCA"), title = "Madagascar" )
gridExtra::grid.arrange( g1, g2, ncol = 2)

pdf(width = 10, height = 20, file = "missing_RCA.pdf"); print(g1); dev.off()
pdf(width = 10, height = 20, file = "missing_Mada.pdf"); print(g2); dev.off()

mp = finalfit::missing_glimpse( datt ) %>% 
  dplyr::select( label, missing_percent ) %>% 
  dplyr::mutate( missing_percent = as.numeric( missing_percent)  ) %>%  
  dplyr::arrange( -missing_percent) %>%
  dplyr::rename( varible = label, percent_missing = missing_percent )

 tibble::as_tibble(mp) %>% print( n = Inf)
```

## Potential Outliers (Immuno/Cyto variables)

Before considering potential associations (using correlation plots or regression), we first look for potential outliers. The values below are greater than twice the 99th percentile of each dataset. The `id = upper99` shows the 99th percentile for each variable. Of note: the immunoglobulin data in the feces seems to be pretty variable and much less stable than in the blood.

```{r}
#-- potential outliers
rbind.data.frame( high, bads ) %>% knitr::kable()
write.csv( bads, file = "potential_outliers.csv")

#-- visualization outliers
gg = ggplot( long %>% dplyr::filter( name %in% name[grepl( "igg", name)] )  ) +
  geom_col(aes(x = value, y = name, group = id, col = flagged)
           , position = "dodge" )+
  # facet_wrap(. ~ flagged)+
  theme_minimal()+
  # scale_x_continuous( limits =c(0,40e3)) +
  theme( legend.position = "none")

###-- to save as a PDF, uncomment the pdf() and dev.off() lines
 pdf( width = 10, height = 10, file = "outlier_plot.pdf")
print(gg)
dev.off()
```

The following plot shows the first two principle components of the `r nrow(immu_pcoa)` subjects who were not missing any of the fecal or blood immunoglobulin or cytokine observations. 

```{r}
gg = ggplot( immu_pcoa) + 
  geom_point(aes(y = PC2, x = PC1, col = flagged, shape = pays ))+
  # facet_grid(. ~ pays )+
  theme_minimal()+
  labs( col = "Potential Outlier", shape = "Country")
pdf( width = 3, height = 5, file = "outlier_pcoa.pdf")
print(gg)
dev.off()
```

## PC plots of cytokine variables


What if we look only at the cytokines in the blood? These data is only available for Madagascar and there seems to be no major shift in the overall profile of cytokines 

```{r}
vars  =  blood_cytos 
prefix = gsub(pattern = "_blood", "", vars) %>% unique()
small = datt %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("cyto","comp") ) %>%
    unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(cyto ) ) %>%
  make_pca_vars( amp_vars = prefix, normer = NULL)

gg = ggplot(small) + 
  geom_point(aes(y = PC2, x = PC1, col = haz, shape = pays )
             , alpha = 0.5 )+
  facet_grid(. ~ comp ) +
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small$id),")")
        , subtitle = "Blood Cytokine Variables")

pdf( width = 5, height = 5, file = "Bloodcytokinesstunted.pdf")
print(gg)
dev.off()

```

Look at immunoglobulins in all three compartments and then in compartments two-by-two

```{r}
immus_red <- c('igg1_blood', 'igg1_duo', 'igg1_feces', 'igg2_blood', 'igg2_duo', 'igg2_feces', 'igg3_blood', 'igg3_duo', 'igg3_feces', 'igg4_blood', 'igg4_duo', 'igg4_feces', 'igm_blood', 'igm_duo', 'igm_feces', 'iga_blood', 'iga_duo','iga_feces')

vars   = immus_red
prefix = gsub(pattern = "_feces|_blood|_duo", "", vars) %>% unique()

small = datt %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red) ) %>%
  make_pca_vars( amp_vars = prefix, normer = NULL)

gg = ggplot(small) + 
  geom_point(aes(y = PC2, x = PC1, col = pays, shape = haz )
             , alpha = 0.5 )+
  facet_grid(. ~ comp ) +
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small$id),")")
        , subtitle = "Blood, duodenal and fecal Immunoglobulin Variables")

print(gg)

gg2 = ggplot(small) + 
  geom_point(aes(y = PC2, x = PC1, col = pays, shape = comp )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small$id),")")
        , subtitle = "Comparison of blood, duodenal and fecal Immunoglobulin Variables")

print(gg2)

# now take out the outliers and plot again

datt2 <- subset(datt, (datt$id%in% bad_ids)==FALSE)

nrow(datt)
nrow(datt2)

small2 = datt2 %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red) ) %>%
  make_pca_vars( amp_vars = prefix, normer = NULL)


gg3 = ggplot(small2) + 
  geom_point(aes(y = PC2, x = PC1, col = comp, shape = pays )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small2$id),")")
        , subtitle = "Comparison of blood, duodenal and fecal Immunoglobulin Variables")

pdf( width = 5, height = 5, file = "Immunoglobulinsbycompartments.pdf")
print(gg3)
dev.off()

immus_red_noduod=c('igg1_blood',  'igg1_feces', 'igg2_blood', 'igg2_feces', 'igg3_blood', 'igg3_feces', 'igg4_blood',  'igg4_feces', 'igm_blood',  'igm_feces', 'iga_blood', 'iga_feces')

vars   = immus_red_noduod
prefix2 = gsub(pattern = "_feces|_blood", "", vars) %>% unique()


small3 = datt2 %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red_noduod","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red_noduod) ) %>%
  make_pca_vars( amp_vars = prefix2, normer = NULL)

gg4 = ggplot(small3) + 
  geom_point(aes(y = PC2, x = PC1, col = pays, shape = comp )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small3$id),")")
        , subtitle = "Comparison of blood and fecal Immunoglobulin Variables")

pdf( width = 5, height = 5, file = "BloodFecesImmunoglobulinsPCA.pdf")
print(gg4)
dev.off()

immus_red_fecesduod=c('igg1_duo',  'igg1_feces', 'igg2_duo', 'igg2_feces', 'igg3_duo', 'igg3_feces', 'igg4_duo',  'igg4_feces', 'igm_duo',  'igm_feces', 'iga_duo', 'iga_feces')

vars   = immus_red_fecesduod
prefix2 = gsub(pattern = "_feces|_duo", "", vars) %>% unique()


small4 = datt2 %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red_fecesduod","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red_fecesduod) ) %>%
  make_pca_vars( amp_vars = prefix2, normer = NULL)

gg5 = ggplot(small4) + 
  geom_point(aes(y = PC2, x = PC1, col = comp, shape = pays )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small4$id),")")
        , subtitle = "Comparison of blood and duodenal Immunoglobulin Variables")

pdf( width = 5, height = 5, file = "DuodenumFecesImmunoglobulinsPCA.pdf")
print(gg5)
dev.off()

```
Look at immunoglobulins in each compartment and stunting

```{r}
immus_red_blood <- c('igg1_blood',  'igg2_blood',  'igg3_blood',  'igg4_blood', 'igm_blood',  'iga_blood')


vars   = immus_red_blood
prefix2 = gsub(pattern = "_blood", "", vars) %>% unique()


small3 = datt2 %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red_blood","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red_blood) ) %>%
  make_pca_vars( amp_vars = prefix2, normer = NULL)

gg4 = ggplot(small3) + 
  geom_point(aes(y = PC2, x = PC1, col = haz, shape = pays )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small3$id),")")
        , subtitle = "Comparison of blood Immunoglobulin Variables according to stunting")

pdf( width = 5, height = 5, file = "BloodImmunoglobulinsPCAstunted.pdf")
print(gg4)
dev.off()

immus_red_feces=c( 'igg1_feces', 'igg2_feces',  'igg3_feces',   'igg4_feces',   'igm_feces',  'iga_feces')

vars   = immus_red_feces
prefix2 = gsub(pattern = "_feces", vars) %>% unique()


small4 = datt2 %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red_feces","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red_feces) ) %>%
  make_pca_vars( amp_vars = prefix2, normer = NULL)

gg5 = ggplot(small4) + 
  geom_point(aes(y = PC2, x = PC1, col = haz, shape = pays )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small4$id),")")
        , subtitle = "Fecal Immunoglobulin Variables by stunting")

pdf( width = 5, height = 5, file = "FecesImmunoglobulinsPCAstunted.pdf")
print(gg5)
dev.off()

immus_red_duo=c( 'igg1_duo', 'igg2_duo',  'igg3_duo',   'igg4_duo',   'igm_duo',  'iga_duo')

vars   = immus_red_duo
prefix2 = gsub(pattern = "_duo", vars) %>% unique()


small4 = datt2 %>% 
  dplyr::select(id, vars, pays, haz ) %>% 
  non_missing( controls = vars )  %>% 
  tidyr::pivot_longer( cols = vars) %>%
  tidyr::separate( col = "name", sep = "_", into = c("immus_red_duo","comp") ) %>%
  unique() %>%
  tidyr::pivot_wider( id_cols = c(id, pays, haz , comp )
                      , values_from = value, names_from = c(immus_red_duo) ) %>%
  make_pca_vars( amp_vars = prefix2, normer = NULL)

gg5 = ggplot(small4) + 
  geom_point(aes(y = PC2, x = PC1, col = haz, shape = pays )
             , alpha = 0.5 )+
  theme_minimal() +
  scale_colour_viridis_d( end = .5, direction = -1, option = "C") + 
  labs( title = paste("Principle Components (n =", lunique(small4$id),")")
        , subtitle = "Duodenal Immunoglobulin Variables by stunting")

pdf( width = 5, height = 5, file = "DuodenumImmunoglobulinsPCAstunted.pdf")
print(gg5)
dev.off()

```

```

## Associations among Biomarkers (Immuno, Cytokines) 

Visualization of two-way associations (among the same biomarker, across compartments) using Spearman correlations. To better see the associations, we remove the potential outliers flagged earlier. 

After adjusting the p-values for multiple comparisons (within a compartment), none of the associations is statistically significant.

```{r, fig.height = 50, fig.width = 8}
prefix = c(immus, blood_cytos) %>% 
  lapply( FUN = function(x) strsplit(x, split = "_")[[1]][1]) %>% 
  unlist() %>% 
  unique() %>% 
  sort()

k = 0; out = NULL; gg =list()
for( j in prefix){
for( comp in c("blood","duo")){
  # j = prefix[24]; comp = "duo"
  temp = immu_dat %>% 
    dplyr::select(id, dplyr::contains( j ), haz ) %>%
    # dplyr::mutate( flagged = (id %in% bad_ids) )
    dplyr::filter( !id %in% bad_ids )
  # colnames(temp)
 
  
  ycol = paste0( j, "_feces")
  xcol = paste0( j, "_", comp)

  if( sum( c(ycol, xcol) %in% colnames(temp) ) == 2 ){   
    k = k + 1 
    
  varr = paste0(j, "_", c( comp ,"feces"))
  smm = temp[, varr ] %>% non_missing( controls = varr) %>% as.matrix()
 
  smm = cbind.data.frame( pval = cor.test( smm[,1], smm[,2]
                          , method = "spearman")$p.value
                          , name = j
                          , comp = comp )
  out = rbind.data.frame( out, smm )
  
   gg[[k]] = ggpubr::ggscatter( temp
                     , x =  xcol
                     , y =  ycol
                     , add = "reg.line"
                     , cor.method = "spearman"
                     , conf.int = TRUE
                     , cor.coef = TRUE 
                     , color = "grey44") + 
   # gg[[k]] =  ggplot( temp ) +
   #    geom_point(aes_string( y = ycol, x = xcol
   #                           # , col = "flagged"
   #                           ), alpha = 0.5 )+
      scale_color_viridis_d( end = .9, direction = -1) + 
      theme_minimal()+ 
      theme(legend.position = "none")
  }# end plot
}# end loop
}# end loop

out %>%
  dplyr::group_by( comp) %>% 
  dplyr::mutate(padj = p.adjust( pval, method = "BH")) %>%
  dplyr::arrange( padj ) %>% 
  dplyr::filter( padj < .10 ) %>% 
  print(n = Inf) 

p = 2
gridExtra::marrangeGrob( gg , ncol = p, nrow = 5)
```

## Relationship between Stunting and AAT, Calprotectine, Ferritine, CRP, Citrulline, Igf
## Stunting and Inflammation

## Binary associations between Cytokines, Immunoglobulins and stunting

The following tables are organized by country of origin. 

**Note on p-values** For categorical or binary variables (such as the pathogen presence/absence variables), p-values are from a Chi-Squared test of independence. For continuous variables, p-values are from a Wilcoxon rank sum test. This analysis does not consider any controls (such as sex or age).

```{r, results = "asis"}

non_duo_immus_red = c( 'iga_feces', 'iga_blood', 'ige_blood', 'igg1_blood', 'igg1_feces', 'igg2_blood', 'igg2_feces', 'igg3_blood', 'igg3_feces', 'igg4_blood', 'igg4_feces', 'igm_blood', 'igm_feces')

form = paste("~ "
  , paste0( non_duo_immus_red, collapse = " + "), "| stunted") %>% 
  as.formula()

for( country in c("RCA","Madagascar")){# country = "RCA"
gg = table1::table1( data = datt %>% dplyr::filter( pays == country ) 
           , form 
           , overall = FALSE
           , extra.col = list( `P-value` = pvalue_table1 )
           , caption = paste("Data in", country) )
print(gg)
}

#-- for the duodenal variables

form = paste("~ "
  , paste0( duo_immus, collapse = " + "), "| haz") %>% 
  as.formula()

for( country in c("RCA","Madagascar")){
gg = table1::table1( data = datt %>% dplyr::filter( haz != "non-stunted" 
                                                    & pays == country)
       , form 
       , overall = FALSE
       , extra.col = list( `P-value` = pvalue_table1)
       , caption = paste("Data in", country, "(excluding non-stunted children)")
                  )
print(gg)
}# end loop

#-- for the blood cytokines

form = paste("~ "
  , paste0( blood_cytos, collapse = " + "), "| stunted") %>% 
  as.formula()

for( country in c("Madagascar")){# country = "RCA"
gg = table1::table1( data = datt %>% dplyr::filter( pays == country ) 
           , form 
           , overall = FALSE
           , extra.col = list( `P-value` = pvalue_table1 )
           , caption = paste("Data in", country) )
print(gg)
}# end loop

# now control for co-factors for the significantly associated variables

controls = c("age","sex","ige_blood","il2r_blood", "vegf_blood", "mig_blood", "pays")
fec = datt  %>% non_missing( controls = controls )

mod = lm( haz_cont ~ age + sex + ige_blood+ il2r_blood+ vegf_blood+ mig_blood
          , data = fec )
nice_print( mod )

many_interactions( mod = mod
                   , elas_data = fec
                   , pred_x = c("ige_blood", "il2r_blood", "vegf_blood", "mig_blood")
                   , cat_z = "sex")

```


**Statistical Analysis** Here we fit a linear model to the height-for-age z-score on fecal samples, where the covariates are the inflammation variables (AAT_feces, cal_feces), age, sex, and country of origin. 

**Initial Findings** No effects are significant except for citrulline


```{r, results = "asis"}
controls = c("age","sex","pays","AAT_feces","cal_feces", "citrulline", "pays", "valin", "alanin", "isoleucin", "leucin")
fec = datt  %>% non_missing( controls = controls )

mod = lm( haz_cont ~ age + sex + pays *( log10(AAT_feces) + log10(cal_feces) + citrulline + valin + alanin + isoleucin + leucin )
          , data = fec )
nice_print( mod )

many_interactions( mod = mod
                   , elas_data = fec
                   , pred_x = c("AAT_feces", "cal_feces", "citrulline", "alanine", "leucin", "isoleucin", "valin" )
                   , cat_z = "sex")
```

now we also control taking into account anemia

```{r, results = "asis"}
controls = c("age","sex","pays","AAT_feces","cal_feces","haz", "anemia", "valin", "alanine", "isoleucin", "leucin")
fec = datt  %>% non_missing( controls = controls )

mod = lm( haz_cont ~ age + sex + anemia + pays *( log10(AAT_feces) + log10(cal_feces) + crp + citrulline + ferritine +valin + alanine + isoleucin + leucin)
          , data = fec )
nice_print( mod )

many_interactions( mod = mod
                   , elas_data = fec
                   , pred_x = c("AAT_feces", "cal_feces", "crp", "igf_n", "citrulline", "ferritine", "alanine", "leucin", "isoleucin", "valin" )
                   , cat_z = "anemia")
```

## Differences in Stunting across Biomarkers (Immuno, Cytokines) 

**Statistical Approach** Next we examine linear models with a few standard controls (age, sex, country of origin, season, sda and anemia) and look at the associations (and p-values) with each biomarker, one-at-a-time. We adjust the p-values for multiple comparisons using Benjamin-Hochberg. This analysis will help us see what associations are still worth exploring. 

Here we consider the immunoglobulin and cytokine variables.

**Initial Findings** Among the biomarkers, only ige_blood, VEGF and IL2r have (unadjusted) p-values less than 0.05 but only IgE remains significant after multiple testing. 

```{r, eval = TRUE }

controls = c( "age", "sex", "sda", "saison"
              , "hemoglobine")
ttt  = NULL

igf = c("igf_n")

varrs = unique( c(immus, blood_cytos, igf) )

for ( varr in varrs ){ 

#-- filtering to non-missing
small = datt %>% non_missing( c(controls, varr))
#-- in case country is all the same
bad = ifelse( (small$pays %>% unique() %>% length() )  <  2
              , " + ", " + pays + ")

form = formula( paste0( "haz_cont ~ "
                        , paste0( controls, collapse = " + ")
                        , bad , varr ))
mod = lm( data = small, formula = form )
tt  = mod %>% 
  summary %$% 
  coefficients[ , c('Estimate','Pr(>|t|)')] 
tt = tt[ nrow(tt), ] %>% 
  t() %>% 
  data.frame() %>% 
  dplyr::mutate( variable = varr ) %>%
  dplyr::rename( pval = Pr...t.. )

ttt = rbind.data.frame( ttt, tt )
 #summary(mod)
}

#nice_print( mod ); summary(mod)
#--- from here, it seems only ige_blood is significantly associated
ttt %>% 
  dplyr::mutate( padj = stats::p.adjust(p = pval, method = "BH")
                 , sig = padj < 0.05 
                 , Estimate = signif( Estimate, 3) ) %>%
  dplyr::arrange( pval, -sig ) %>%
  # dplyr::arrange( - sig ) %>% 
  tibble::as_tibble() %>% 
  print( n = Inf)

```

The scatterplots between those variables and `haz_cont` suggest that the association with `IgE_blood`may have a meaningful association. 

```{r}

#-- worth exploring (unadjusted p-value < 0.05 )
these = ttt[ttt$pval < 0.05, "variable"] %>% unlist()
gg = list(); hh = list(); k = 0

for(j in these){ # j = these[1]
  k = k + 1
  
  temp = datt %>% 
    dplyr::select( id, sex, pays, all_of(j), haz_cont, stunted, haz ) %>%
    dplyr::mutate( flagged = (id %in% bad_ids))
  
  gg[[k]] = ggplot( temp ) + 
    geom_point(aes_string(x = j, y = "haz_cont", col = "pays")) +
    geom_smooth( aes_string(x = j, y = "haz_cont")
                 , method = "lm", alpha = 0.5) +
    theme_minimal()+
    theme( legend.position = "top")
  
  hh[[k]] = ggplot( temp ) +
    geom_boxplot( aes_string(y = paste0("sqrt(",j ,")")
                           , x = "haz", fill = "haz")) +
    theme_minimal()+ 
    theme( legend.position = "none")
}#end loop

gridExtra::marrangeGrob( hh, ncol = 4, nrow = 1)

gridExtra::marrangeGrob( gg, ncol = 4, nrow = 1)

```


**Statistical Approach** Next we construct bootstrapped datasets and regress `haz_cont` on the same controls along with biomarkers from the same compartment (e.g. blood, feces, duodenal). We also repeat the analysis with the potential outliers and without them. 

Because there are only 64 observations with all of the blood cytokine variables measured, we make two models, one with all and one having instead only consider IgA, IgE, and Igg1 through Igg4, IgM, which have more than 600 available observations. 

**Initial Analysis  **

With the blood variables, only`ige_blood` is statistically significant, with higher values associated with lower `haz_cont` scores. This finding is consistent whether or not the outliers are removed.

A bootstrap analysis suggests that increases in `ige_blood` is significantly associated with lower `haz_cont`, controlling for the other controls and fecal variables. 



```{r, results = "asis" }

set.seed(81053)

controls = c( "age", "sex", "sda", "saison"
              , "hemoglobine")

for( comp in c("feces", "duo", "blood")){ # comp = "blood"; comp = "duo"; comp = "feces"

  
##--- linear regression (compartment specific)
immu_controls =  immus[ grepl( comp , immus )]

if (comp != "feces"){ # duodenal and blood have only values in one country
  interact = "+ ("
} else interact = "+ pays + ("

all_controls = c(controls, immu_controls )
form  = formula( paste(
  c("haz_cont ~ "
    , paste0( controls, collapse = "+")
                 , interact 
                 , paste0( immu_controls, collapse = "+")
                 , ")"), collapse  = " ") )

for (include_outlier in c(FALSE, TRUE) ) { # include_outlier = TRUE

small = dat %>% non_missing( controls = all_controls) %>%
  dplyr::select( all_of( c("haz_cont", all_controls, "pays", "id" ) ))

if (include_outlier == FALSE) small = small %>% dplyr::filter( !id %in% bads$id)

mod   = lm( formula = form, data = small)
print( paste("including outliers:", include_outlier ) )
nice_print( mod )
# mod %>% summary() %>% print()

}# end outlier

#-- corrplot
tt = small %>% dplyr::select( all_of( immu_controls ) ) %>% cor( method = "spearman")
corrplot::corrplot( tt )

#--- bootstrapping
pint = bootstrap_coef( small = small, form = form, times = 2000 )
gg   = plot_bootstrap( pint )
print(gg)
}

```

## Association between AAT and Calprotectine in different compartments

**Statistical Approach** We regress each biomarker in the feces with its counterpart in the duodenum, controlling for age, sex, and country of origin. We allow for an interaction effect with country of origin. To aim for residuals that are normally distributed, we take a square-root transformation of AAT, and we take a logarithm-transformation for calprotectine.

**Initial Findings** There is no association between calprotectine measured in the feces and calprotectine measured in the duodenum (p-value > 0.10), whether we use no transformation, a square-root, or a logarithm transformation.

Depending on the transformation, there either is or is not a statistically significant association between AAT measured in the feces and in the duodenum. What makes this analysis challenging is that nearly 33% of the AAT_duodenal values are at or below the detection limit. 

If we take these values as they are and use the original scale, there is a statistically significant association (that differs between countries) between AAT in the feces and AAT in the duodenum. However, if we take a logarithm or square-root transformation, the effect is no longer significant (p-values > 0.05). 

The Spearman correlation between AAT measured in the feces and AAT measured in the duodenum is not significant (p-value = 0.82). 


```{r aat, results = "asis"}
for( biomark in c("AAT","calprotectine")){ # biomark = "AAT"
  
  if ( biomark == "AAT" ) { 
    controls = c("AAT_duo","AAT_feces")
    form = formula(  sqrt(AAT_feces) ~ pays* (sex + age +  sqrt(AAT_duo)))
    yvar = "AAT_feces"
    xvar = "AAT_duo"
  } else {
     controls = c("cal_duo","cal_feces")
    form = formula( log10(cal_feces) ~ pays* (sex + age +  log10(cal_duo)))
    yvar = "cal_feces"
    xvar = "cal_duo"
    
  }
  
  
small = datt %>% dplyr::select( id, pays, sex
                                , stunted, AAT_duo, AAT_feces, age, anemia
                                , cal_feces, cal_duo ) %>%
  dplyr::filter( AAT_duo > 0 ) %>%
  non_missing( controls = controls  ) %>%
  dplyr::mutate( ageyears = dplyr::case_when( age <= 36 ~ "(0-3]"
                                            , age <= 48 ~ "(3-4]"
                                            , age > 48 ~ "(4-6]")) %>%
  dplyr::mutate( aat_small = ( AAT_duo <= 0.012))

# table(small$pays)
mod = lm( formula = form, data = small)
nice_print(mod) # summary(mod)
 # print( summary(mod) ) ; residuals( mod) %>% hist(); qqnorm( residuals(mod))
pint = bootstrap_coef( small = small, form = form )
plot_bootstrap( pint ) +
  labs( title = paste("Bootstrap Analysis of", biomark) )

gg = ggpubr::ggscatter( data = small
                   , x = xvar, y = yvar
                   , cor.method = "spearman"
                   , add = "reg.line"
                     , conf.int = TRUE
                     , cor.coef = TRUE)

print(gg)
}# end loop

```


## Complete-Case Analysis (AAT, Citrulline, Hemoglobine, etc.)

To check if the statistical significant effects appear in the dataset, we consider fitting a linear model with only the complete-case data. 

As shown in the table of non-missing observations (below), of the original 806 observations, 608 are non-missing for AAT_feces, citrulline, ferritine, and hemoglobine and 583 for AAT_feces, cal_feces, citrulline, crp, ferritine, hemoglobine

```{r}
controls = c("citrulline","hemoglobine","AAT_feces"
             ,"ferritine", "crp", "cal_feces", "valin", "leucin", "isoleucin")
missing_size( dat, controls = controls ) %>% print( n = Inf)

controls2 = c("citrulline","hemoglobine","AAT_feces"
             ,"ferritine", "crp", "cal_feces", "igf_n", "valin", "leucin", "isoleucin")
missing_size( dat, controls = controls2 ) %>% print( n = Inf)


```

## Associations between Biomarkers as well as with stunting

**Statistical Approach** We consider the biomarkers: citrulline, fertitine, crp, and logarithm-transformed AAT_feces (`AATmggdePS`) and cal_feces (`CALPROTECTINEggdePS`). 
In a seperate model, we consider in addition also Igf.

We regress each biomarker on the controls: age, sex, sda, season, cough, runny nose, stuffy nose,  country of origin, hemoglobin and the remaining biomarkers. To test for robustness, we consider 1000 bootstrap datasets. In this analysis, we do not include an interaction with country of origin.  

### Initial Findings (Biomarkers and Stunting)

We present the linear regression and the plots of predicted values across biomarkers excluding AA ramefiés


```{r, results = "asis", fig.width = 10, fig.height = 13 }
##----
controls = c( "age", "sex", "sda", "saison", "hemoglobine")
biomarks = c("haz_cont", "ferritine", "citrulline"
             ,"crp","log_AAT","log_cal", "ige_blood" )

small = dat %>% 
  non_missing( controls = c("cal_feces", "crp", "ferritine"
                            ,"citrulline","AAT_feces")) %>%
  dplyr::mutate( log_cal = log10( cal_feces)
                 , log_AAT = log10( AAT_feces ) )

set.seed(83150832)
outt = NULL 

bootplot = list()
i = 0
for (biomark in biomarks){ 
  i = i + 1 
  # biomark = biomarks[1]; biomark = "log_AAT"; biomark = "log_cal"
  
  if( biomark %in% c("log_AAT" ,"log_cal", "ige_blood")){
    controlz = c(controls, c("ascaris","trichuris"
                             , "num_genes"
    ) ) 
  } else{
    controlz = controls 
  }
  
  tiny = small %>% non_missing( controls = controlz )
  
  others = setdiff( biomarks, biomark)
  form   = formula( paste0( biomark, "~"
                            , paste0( controlz, collapse = "+")
                            # , "+ pays * ("
                            , "+ pays + ("
                            , paste0( others, collapse = "+")
                            , ")"))
  
  modd = lm( formula = form, data = tiny )
  nice_print( modd )
  # print( biomark )
  # print( summary(modd))
  
  ##-- storing p-values 
  out = (modd %>% summary() %$% coefficients)[, 4] %>% 
    data.frame() %>%
    tibble::rownames_to_column( var = "term") %>%
    dplyr::mutate( y = biomark )
  
  outt = rbind.data.frame( outt, out )
  write.csv(outt, "associations.csv")
  
  #--- bootstrapping 
  pint = bootstrap_coef(small = tiny, form = form, times = 1000)
  bootplot[[i]]   = plot_bootstrap( pint ) +
    labs(title = paste("Bootstrap Estimates\nRegression on", biomark)
         , subtitle = paste("Sample size =", nrow(tiny)))
  
  ##--- plot predictions as a panel.
  gg = many_interactions(mod = modd, elas_data = tiny
                         , pred_x = others # multiple ones
                         , cat_z = "saison" )
  
  #-- as a panel
  do.call( gridExtra::grid.arrange, gg)
  
  
  
}# end biomarker loop
```

### Sensitivity Test 

Here are the bootstrap plots of significant variables. If a variable is statistically significant in the bootstrap samples, but not in the full analysis, we do not count it as statistically significant. 

```{r, fig.height = 6, fig.width = 7 }
bootplot 
```


We present the linear regression and the plots of predicted values across biomarkers including AA ramefiés


```{r, results = "asis", fig.width = 10, fig.height = 13 }
##----
controls = c( "age", "sex", "sda", "saison",  "hemoglobine")
biomarks = c("haz_cont", "ferritine", "citrulline", "bcaa_geomean"
             ,"crp","log_AAT","log_cal", "ige_blood" )

small = dat %>% 
  non_missing( controls = c("cal_feces", "crp", "ferritine"
       ,"citrulline","AAT_feces")) %>%
  dplyr::mutate( log_cal = log10( cal_feces)
               , log_AAT = log10( AAT_feces ) )

set.seed(83150832)
outt = NULL 

bootplot = list()
i = 0
for (biomark in biomarks){ 
i = i + 1 
  # biomark = biomarks[1]; biomark = "log_AAT"; biomark = "log_cal"

  if( biomark %in% c("log_AAT" ,"log_cal", "ige_blood")){
    controlz = c(controls, c("ascaris","trichuris"
                              , "num_genes"
                             ) ) 
  } else{
    controlz = controls 
  }

  tiny = small %>% non_missing( controls = controlz )

  others = setdiff( biomarks, biomark)
  form   = formula( paste0( biomark, "~"
                            , paste0( controlz, collapse = "+")
                            # , "+ pays * ("
                               , "+ pays + ("
                            , paste0( others, collapse = "+")
                            , ")"))
  
  modd = lm( formula = form, data = tiny )
  nice_print( modd )
  # print( biomark )
  # print( summary(modd))
  
  ##-- storing p-values 
  out = (modd %>% summary() %$% coefficients)[, 4] %>% 
    data.frame() %>%
    tibble::rownames_to_column( var = "term") %>%
    dplyr::mutate( y = biomark )
  
  outt = rbind.data.frame( outt, out )
  write.csv(outt, "associations.csv")

  #--- bootstrapping 
pint = bootstrap_coef(small = tiny, form = form, times = 1000)
bootplot[[i]]   = plot_bootstrap( pint ) +
  labs(title = paste("Bootstrap Estimates\nRegression on", biomark)
       , subtitle = paste("Sample size =", nrow(tiny)))

##--- plot predictions as a panel.
gg = many_interactions(mod = modd, elas_data = tiny
                  , pred_x = others # multiple ones
                  , cat_z = "saison" )

#-- as a panel
do.call( gridExtra::grid.arrange, gg)



}# end biomarker loop
```

### Sensitivity Test 

Here are the bootstrap plots of significant variables. If a variable is statistically significant in the bootstrap samples, but not in the full analysis, we do not count it as statistically significant. 

```{r, fig.height = 6, fig.width = 7 }
bootplot 
```

We present the linear regression and the plots of predicted values across biomarkers including also Igf and ramefied AA


```{r, results = "asis", fig.width = 10, fig.height = 13 }
##----
controls = c( "age", "sex", "sda", "saison", "hemoglobine")
biomarks = c("haz_cont", "ferritine", "citrulline", "bcaa_geomean"
             ,"crp","log_AAT","log_cal", "igf_n" )

small = dat %>% 
  non_missing( controls = c("haz_cont", "ferritine", "citrulline", "bcaa_geomean"
             ,"crp","log_AAT","log_cal", "igf_n" )) %>%
  dplyr::mutate( log_cal = log10( cal_feces)
                 , log_AAT = log10( AAT_feces ) )

set.seed(83150832)
outt = NULL 

colnames(small)

bootplot = list()
i = 0
for (biomark in biomarks){ 
  i = i + 1 
   #biomark = biomarks[1]; biomark = "log_AAT"; biomark = "log_cal"
  
  if( biomark %in% c("log_AAT" ,"log_cal", "ige_blood")){
    controlz = c(controls, c( "num_genes"
    ) ) 
  } else{
    controlz = controls 
  }
  
  tiny = small %>% non_missing( controls = controlz )
  
  others = setdiff( biomarks, biomark)
  form   = formula( paste0( biomark, "~"
                            , paste0( controlz, collapse = "+")
                            # , "+ pays * ("
                            , "+ ("
                            , paste0( others, collapse = "+")
                            , ")"))
  
  modd = lm( formula = form, data = tiny )
  nice_print( modd )
  # print( biomark )
  # print( summary(modd))
  
  ##-- storing p-values 
  out = (modd %>% summary() %$% coefficients)[, 4] %>% 
    data.frame() %>%
    tibble::rownames_to_column( var = "term") %>%
    dplyr::mutate( y = biomark )
  
  outt = rbind.data.frame( outt, out )
  write.csv(outt, "associations2.csv")
  
  #--- bootstrapping 
  pint = bootstrap_coef(small = tiny, form = form, times = 1000)
  bootplot[[i]]   = plot_bootstrap( pint ) +
    labs(title = paste("Bootstrap Estimates\nRegression on", biomark)
         , subtitle = paste("Sample size =", nrow(tiny)))
  
  ##--- plot predictions as a panel.
  gg = many_interactions(mod = modd, elas_data = tiny
                         , pred_x = others # multiple ones
                         , cat_z = "saison" )
  
  #-- as a panel
  do.call( gridExtra::grid.arrange, gg)
  
  
  
}# end biomarker loop
```

### Sensitivity Test 

Here are the bootstrap plots of significant variables. If a variable is statistically significant in the bootstrap samples, but not in the full analysis, we do not count it as statistically significant. 

```{r, fig.height = 6, fig.width = 7 }
bootplot 
```


### Association between biomarkers and hemoglobin levels 


**Statistical Approach** Next we examine linear models with a few standard controls (age, sex, country of origin, season, sda) and look at the associations (and p-values) with each biomarker, one-at-a-time. We adjust the p-values for multiple comparisons using Benjamin-Hochberg. This analysis will help us see what associations are still worth exploring. 


**Initial Findings** 

```{r, eval = TRUE }

controls = c( "age", "sex", "sda", "saison"
              , "stunted")
ttt  = NULL

varrs = c("citrulline","bcaa_geomean", "AAT_feces" ,"ferritine", "crp", "cal_feces", "ige_blood")

for ( varr in varrs ){ 
  
  #-- filtering to non-missing
  small = datt %>% non_missing( c(controls, varr))
  #-- in case country is all the same
  bad = ifelse( (small$pays %>% unique() %>% length() )  <  2
                , " + ", " + pays + ")
  
  form = formula( paste0( "hemoglobine ~ "
                          , paste0( controls, collapse = " + ")
                          , bad , varr ))
  mod = lm( data = small, formula = form )
  tt  = mod %>% 
    summary %$% 
    coefficients[ , c('Estimate','Pr(>|t|)')] 
  tt = tt[ nrow(tt), ] %>% 
    t() %>% 
    data.frame() %>% 
    dplyr::mutate( variable = varr ) %>%
    dplyr::rename( pval = Pr...t.. )
  
  ttt = rbind.data.frame( ttt, tt )
  #summary(mod)
}

#nice_print( mod ); summary(mod)

ttt %>% 
  dplyr::mutate( padj = stats::p.adjust(p = pval, method = "BH")
                 , sig = padj < 0.05 
                 , Estimate = signif( Estimate, 3) ) %>%
  dplyr::arrange( pval, -sig ) %>%
  # dplyr::arrange( - sig ) %>% 
  tibble::as_tibble() %>% 
  print( n = Inf)

```

The scatterplots between those variables and `hemoglobine` suggest that the association with all but AAT_feces and IgE may have a meaningful association. 

```{r}

#-- worth exploring (unadjusted p-value < 0.05 )
these = ttt[ttt$pval < 0.05, "variable"] %>% unlist()
gg = list(); hh = list(); k = 0

for(j in these){ # j = these[1]
  k = k + 1
  
  gg[[k]] = ggplot( datt ) + 
    geom_point(aes_string(x = j, y = "hemoglobine", col = "pays")) +
    geom_smooth( aes_string(x = j, y = "hemoglobine")
                 , method = "lm", alpha = 0.5) +
    theme_minimal()+
    theme( legend.position = "top")
  
  }#end loop


gridExtra::marrangeGrob( gg, ncol = 5, nrow = 1)
```
